Material Culture Networks in Tonto Basin

### 2023-01-12 manuscript.Rmd
minor revisions

---
title: Material Culture Networks in Tonto Basin
output:
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
  rticles::tf_article: null
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities. This study uses architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data in a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men producing and exchanging projectile points and women producing and exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
always_allow_html: true
---

@@

Network science has many applications for archaeologists, and it can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society, but certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks. Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This is important because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help give better context to these studies. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

@@

## Tonto Basin

Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview]. The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p. 129-133]. The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian. All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.

@@

One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [c.f. @Oliver2001-tu; @Wood2000-ze]. This migration into the Tonto Basin was part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between communities in the Tonto Basin [e.g., @Sliva2002-oz; @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

@@

## Defining Networks and Relations

@@

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj] but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

I used ranked links to keep consistent values between networks. This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity. The top **n** connections between each site were kept and the rest were discarded. In practice, not every node will have the same number of connections. Some may have fewer ties if there are not enough nodes that are similar. More ties can exist when there are ties in the ranks. Because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top **n** of multiple other nodes. In the latter case, this is a good indication of a central node in the network. Because I am using several types of networks, there is no common cutoff value I can use for the number of ties to keep. Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top 3-10 ties. Meaning that each metric is calculated for a network composed of ties with the top 3 connections, and the process is repeated for new networks composed of ties with the top 4 connections, and so forth. Examining the range of these metrics provides a more robust analysis.

@@

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network). Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results. Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality. In this case the centrality measure is simply the mean of the eigenvector centrality for each separate layer, as the layers are not interdependent [@Frost2022-lr]. Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers. These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.

@@

Certain caveats must be acknowledged. All archaeologists deal with missing data, which can greatly affect the results of analysis. Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik]. Regardless of statistical methods, unaccounted for biases will still produce invalid network results [@Bischoff2021-sf]. The most important bias that may affect these findings is chronology. For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq]. Does that invalidate the network analysis? I do not think so. While the networks involve nodes as sites, what the networks are really meant to represent are groups of people. Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation. Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p. 336]. Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location. The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96]. If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on. It is these communities that the network analysis is attempting to capture. Still, evidence indicates that most sites were contemporaneous. For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya]. It is best, though, to think of these interpretations as estimations based on incomplete data.

@@

The architectural components consisted of compounds, roomblocks, and platform mounds. Compounds were the typical residential structures in the area and featured multiple houses built from masonry and/or adobe around a courtyard. Platform mounds were formed from large masonry and adobe retaining walls with a cell-like interior filled with trash and rubble. They were common in the Hohokam region throughout the Classic period. Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku]. Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture. Thus, if a site is labeled a platform mound it may have other architecture. Sites were thus labeled to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.

@@

Figure `r figs` shows the networks with the top five strongest ties between each node. This number of ties represents the fewest ties that connects all nodes in the spatial network. Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network. Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network. Because the network is not complete, one cannot argue that there were no other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location. Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm. The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated. If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location. Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only. Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`). Figure `r figs + 1` provides the eigenvector data by type of network. Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well. The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.

@@

```{r meanEigenVals}
cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
eigenDF %>% 
  filter(name %in% c("ceramics","points")) %>% 
  group_by(name,Architecture) %>% 
  summarize(value = round(mean(value),2)) %>% 
  pivot_wider(names_from = name, values_from = value) %>% 
  ungroup() %>% 
  rowwise() %>% 
  mutate(mean = round(mean(c(ceramics,points)),2)) %>% 
  arrange(desc(mean)) %>%
  # as_hux() %>% 
  # theme_article() %>% 
  # set_caption(cap)
  knitr::kable(caption= cap,
               booktabs = T)
```

@@

# Discusssion {-}

@@

Exchange would be a primary driver in these network dynamics, and clear evidence exists for exchange of several types of material culture. Trade within the basin does not appear to have been dominated by any one group and may have been competitive in nature [@Rice1998-cu, p. 127-128]. Ceramic exchange certainly occurred at a regional level, but even local ceramics were circulated within the basin [@Heidke2000-nr; @Miksa1995-qw; @Stark1992-vu], possibly in return for food [@Clark2004-uw, p.290-291]. Projectile points (either as part of the arrow or separately) may have also been commonly exchanged. There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., @Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob]. Besides Watt's study of knappers, Sliva's [-@Sliva2002-oz, p. 539] analysis of sites from the Roosevelt Community Development Study suggests craft specialization in projectile point production. Mortuary offerings were possibly the intended function for many points [@Sliva2002-oz, p. 543], which may have spurred the increased craft specialization. Shell and stone jewelry, among other artifacts were also produced and exchanged within the basin. Potential loci for exchange are the platform mound sites. The purpose of these sites is still debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p. 112-165; @Doelle1995-qd, p. 439]. If platform mounds were places where exchange between communities regularly took place, then we would expect platform mounds to have higher centrality. Centrality varied among the platform mounds, but on average, platform mounds did have the highest centrality.

@@

There is a potential chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks. While this association is circumstantial and not expected to be uniform across the Hohokam region, roomblocks did have the highest centrality in the ceramic networks. Another reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p. 291]. This second explanation may help explain why roomblock sites had low centrality in the point networks, presuming that points were not exchanged for food. It is beyond this simple analysis to determine the precise reasons, but it is clear that roomblock sites were important for ceramic circulation in Tonto Basin. On the other hand, sites with roomblocks had the lowest centrality in the point networks. This suggests that the influence the occupants of roomblock sites had in pottery networks may not have extended to other spheres of interaction.

@@

Perhaps the reason roomblocks were more central to the ceramic network is because immigrants to the basin had less access to farmland and had to make pottery to get food, as mentioned. This would explain higher centrality for roomblocks and lower centrality for point networks. It does not explain the remaining differences between point and ceramic networks. Projectile points at least were highly gendered, and ceramics probably were as well. The evidence presented here demonstrates that ceramic and point networks vary significantly, and I believe gender likely played an important role.

@@

This is one of the first archaeological applications of multilayer network analysis to consider multiple types of material culture. This is advantageous for studying how types of material culture do or do not co-vary, which aids interpretations of the social interactions that created these patterns. This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin dating primarily from occupations between AD 1275 and 1325. A network was created for each type of data and combined into a multilayer network. Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation were used to study and compare the networks. The findings indicate that the projectile point network was strongly correlated with the spatial network--indicating that sites near each other were more likely to have similar projectile points. Furthermore, sites with roomblocks indicative of immigrants from the north and east of Tonto Basin were, on average, the most highly central sites in the ceramic network; however, they were the least central in the point network. Immigrants may have relied on pottery production to integrate with the local networks, but this relationship did not hold for projectile points. Finally, the results demonstrate major differences between the ceramic and point networks. This finding has major implications for network studies, because many rely on only a single artifact type. If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past. Furthermore, projectile points and ceramics have strong associations with gender. Differences in the ceramic and point networks suggest differences existed between the men's and women's social networks.

This analysis uses a handful of sites from Tonto Basin. Much more data exists and the results discussed here would be greatly strengthened by including more sites and data. What would be perhaps even more useful would be to include sourced obsidian and ceramics. Networks created from artifacts where the origin is known can be used to more strongly infer various types of social interaction. This type of multilayer analysis would also be useful to conduct in other regions. It would be particularly useful for comparative studies to know in what circumstances different types of material culture do or do not co-vary.

@@

Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research. Chris Caseldine's comments greatly improved this paper. All mistakes are my own. The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for contributing the data.

>>>>

---
title: Material Culture Networks in Tonto Basin
output:
  rticles::tf_article: NULL
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit entities. This study uses architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data in a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men producing and exchanging projectile points and women producing and exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
  \usepackage{lineno}
  \linenumbers
csl: american-antiquity.csl
always_allow_html: true
---

@@

Network science has many applications for archaeologists, and it can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society, but certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [e.g., @Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. To name just two: Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks; Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This matters because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help give better context to these studies. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

@@

## Tonto Basin {-}

Tonto Basin is in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview]. The basin was conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p. 129-133]. The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian. All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.

@@

Tonto Basin's position in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions has received much attention [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [@Oliver2001-tu; @Wood2000-ze]. This migration into the Tonto Basin was part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and Salado pottery suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between communities in the Tonto Basin [e.g., @Sliva2002-oz; @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

@@

## Defining Networks and Relations {-}

@@

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I use only the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj] but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

I used ranked links to keep consistent values between networks. This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity. The top **n** connections (**n** varies between 3 and 10) between each site were kept and the rest were discarded. In practice, not every node will have the same number of connections. Some may have fewer ties if there are not enough nodes that are similar. More ties can exist when there are ties in the ranks. Because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top **n** of multiple other nodes. In the latter case, this is a good indication of a central node in the network. Because I am using several types of networks, there is no common cutoff value I can use for the number of ties to keep. Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top 3-10 ties. Meaning that each metric is calculated for a network composed of ties with the top 3 connections, and the process is repeated for new networks composed of ties with the top 4 connections, and so forth. Examining the range of these metrics provides a more robust analysis.

@@

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network). Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R [@RCore2022-a] package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results. Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality. In this case the centrality measure is simply the mean of the eigenvector centrality for each separate layer, as the layers are not interdependent [@Frost2022-lr]. Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers. These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered important variables determining the structure of the ceramic and point networks.

@@

Certain caveats must be acknowledged. All archaeologists deal with missing data, which can greatly affect the results of analysis. Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik]. Regardless of statistical methods, unaccounted for biases will still produce invalid network results [@Bischoff2021-sf]. The most important bias that may affect these findings is chronology. For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq]. Does that invalidate the network analysis? I do not think so. While the networks involve nodes as sites, what the networks are really meant to represent are groups of people. Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation. Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p. 336]. Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location. The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96]. If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on. It is these communities that the network analysis is attempting to capture. Still, evidence indicates that most sites were contemporaneous. For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya]. It is best, though, to think of these interpretations as estimations based on incomplete data and to recognize that these networks are an analytical tool and not a representation of an ancient face-to-face interaction network.

@@

The architectural components consisted of compounds, roomblocks, and platform mounds. Compounds were the typical residential structures in the area and featured multiple houses built from masonry and/or adobe around a courtyard. Platform mounds were formed from large masonry and adobe retaining walls with a cell-like interior filled with trash and rubble. They were common in the Hohokam region throughout the Classic period. Some sites had pit houses dating to earlier occupations, so these few pit houses were removed from the analysis [see @Rice1998-ku]. Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture. Thus, if a site is labeled a platform mound it may have other architecture. Sites were labeled in this manner to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.

@@

Figure `r figs` shows the networks with the top five strongest ties between each node. This number of ties represents the fewest ties that connect all nodes in the spatial network. Fewer ties leave the Cline Mesa sites disconnected from the rest of the network. Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network. Because the network is not complete, one cannot argue that there were no other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location. Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm. The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated. If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location. Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only. Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`). Figure `r figs + 1` provides the eigenvector data by type of network. Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well. The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.

@@

```{r meanEigenVals}
cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
tbls = tbls + 1
eigenDF %>% 
  filter(name %in% c("ceramics","points")) %>% 
  group_by(name,Architecture) %>% 
  summarize(value = round(mean(value),2)) %>% 
  pivot_wider(names_from = name, values_from = value) %>% 
  ungroup() %>% 
  rowwise() %>% 
  mutate(mean = round(mean(c(ceramics,points)),2)) %>% 
  arrange(desc(mean)) %>%
  as_hux() %>%
  theme_article() %>%
  set_caption(cap)
  # knitr::kable(caption= cap,
               # booktabs = T)
```

@@

# Discussion {-}

@@

Exchange would be a primary driver in these network dynamics, and clear evidence exists for exchange of several types of material culture. Trade within the basin does not appear to have been dominated by any one group and may have been competitive in nature [@Rice1998-cu, p. 127-128]. Ceramic exchange certainly occurred at a regional level, but even local ceramics were circulated within the basin [@Heidke2000-nr; @Miksa1995-qw; @Stark1992-vu], possibly in return for food [@Clark2004-uw, p.290-291]. Projectile points (either as part of the arrow or separately) may have also been commonly exchanged. There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., @Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob]. Besides Watt's study of knappers, Sliva's [-@Sliva2002-oz, p. 539] analysis of sites from the Roosevelt Community Development Study suggests craft specialization in projectile point production. Mortuary offerings were possibly the intended function for many points [@Sliva2002-oz, p. 543], which may have spurred the increased craft specialization. Shell and stone jewelry, among other artifacts were also produced and exchanged within the basin. Potential loci for exchange are the platform mound sites. The purpose of these sites is debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p. 112-165; @Doelle1995-qd, p. 439]. If platform mounds were places where exchange between communities regularly took place, then we would expect platform mounds to have higher centrality. Centrality varied among the platform mounds, but on average, platform mounds did have the highest centrality.

@@

There is a potential chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks. While this association is circumstantial and not expected to be uniform across the Hohokam region, roomblocks did have the highest centrality in the ceramic networks. Another proposed reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p. 291]. This second explanation may help explain why roomblock sites had low centrality in the point networks, presuming that points were not exchanged for food. It is beyond this simple analysis to determine the precise reasons, but clearly roomblock sites were important for ceramic circulation. On the other hand, sites with roomblocks had the lowest centrality in the point networks. This suggests that the influence the occupants of roomblock sites had in pottery networks may not have extended to other spheres of interaction.

@@

Perhaps the reason roomblocks were more central to the ceramic network is because immigrants to the basin had less access to farmland and had to make pottery to get food, as mentioned. This would explain higher centrality for roomblocks and lower centrality for point networks. It does not explain the remaining differences between point and ceramic networks. Projectile points at least were highly gendered, and ceramics probably were as well. The evidence presented here demonstrates that ceramic and point networks vary significantly, and gender likely played an important role. More research will be needed to determine how gender roles affected the networks.

@@

This is one of the first archaeological applications of multilayer network analysis to consider multiple types of material culture. This is advantageous for studying how types of material culture do or do not co-vary, which aids interpretations of the social interactions that created these patterns. This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin dating primarily from occupations between AD 1275 and 1325. A network was created for each type of data and combined into a multilayer network. Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation were used to study and compare the networks. The findings indicate that the projectile point network was strongly correlated with the spatial network--indicating that sites near each other were more likely to have similar projectile points. Furthermore, sites with roomblocks indicative of immigrants from the north and east of Tonto Basin were, on average, the most highly central sites in the ceramic network; however, they were the least central in the point network. Immigrants may have relied on pottery production to integrate with the local networks, but this relationship did not hold for projectile points. Finally, the results demonstrate major differences between the ceramic and point networks. This finding has major implications for network studies because many rely on only a single artifact type. If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past. Furthermore, projectile points and ceramics have strong associations with gender. Differences in the ceramic and point networks suggest differences existed between the social networks of men and women.

This analysis used a handful of sites from Tonto Basin. Much more data exists and the results discussed here would be greatly strengthened by including more sites and data. What would be perhaps more useful would be to include sourced obsidian and ceramics. Networks created from artifacts where the origin is known can be used to more strongly infer various types of social interaction. This type of multilayer analysis would also be useful to conduct in other regions. It would be particularly useful for comparative studies to know in what circumstances different types of material culture do or do not co-vary. It is my hope that this study highlights the usefulness of a multilayer network approach utilizing multiple types of material culture.

@@

Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, Michael Brooks, and Ronald Vasquez assisted with this research. Chris Caseldine's comments greatly improved this paper. All mistakes are my own. The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for their hard work.
